---
title: "Ex18 Global proteomics"
output: html_document
---

The global proteomics pipeline uses the R package [PlexedPiper](https://github.com/vladpetyuk/PlexedPiper). It also requires a connection to the DMS to access data packages.


```{r, echo=F}
library(knitr)
knitr::opts_chunk$set(message=F, warning=F)
t0 <- Sys.time(); print(t0)
```


```{r setup}
library(PlexedPiper)
library(dplyr)
library(PNNL.DMS.utils)
library(Biostrings)
library(tidyverse)

data_package_num <- 4135
data_folder <- "data/global_data/"
crosstab_prefix <- "ptrc_ex18"

dir.create(data_folder)

```


# Study design

Study design information in PlexedPiper is encoded in three tables: fractions, samples, and references. 
These tables can be made using metadata and should be stored on the DMS before processing.

```{r read_study_design}
fractions <- read.table("study_design/global_study_design/fractions.txt", sep = "\t", header = T)
samples <- read.table("study_design/global_study_design/samples.txt", sep = "\t", header = T)
references <- read.table("study_design/global_study_design/references.txt", sep = "\t", header = T)

```

# Process MS-GF+ data

MS-GF+ data is processed in several steps. First, read MS-GF+ output from the DMS. (This step can take a while).

```{r read_msgf}
msgf_data_path <- file.path(data_folder, "msgfData_original.RData")
if (file.exists(msgf_data_path)) {
  load(msgf_data_path)
} else {
  msnid <- read_msgf_data_from_DMS(data_package_num)
  save(msnid, file=msgf_data_path)
}
show(msnid)

```


## Correct peak selection and peptide FDR
We use the target-decoy search strategy method described in [(Elias 2010)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2922680/). 
Filtering is done first at peptide level, then at protein level, both with max FDR of 1%.

```{r}
## peak selection & Peptide filter
msnid <- correct_peak_selection(msnid)   ## optional for phospho
msnid <- filter_msgf_data_peptide_level(msnid, 0.01)


```


## Remap accessions

This function remaps UniProt protein accessions to gene symbol.

```{r remap_accessions}
fst_dms_pth <- path_to_FASTA_used_by_DMS(data_package_num)
fst <- readAAStringSet(fst_dms_pth)

conv <- data.frame(NAME = names(fst)) %>%
  mutate(UNIPROT = str_extract(names(fst), "^(\\S)+"),
         SYMBOL = str_extract(names(fst), "GN=(\\S)+")) %>%
  dplyr::select(-NAME) %>%
  mutate(UNIPROT = sub(".*\\|(.*)\\|.*", "\\1", UNIPROT),
         SYMBOL = sub("GN=", "", SYMBOL))

msnid <- remap_accessions_uniprot_to_gene(msnid,
                                          organism_name="Homo sapiens",
                                          conversion_table = conv)

show(msnid)
table(is.na(msnid$accession))
msnid <- apply_filter(msnid, "!is.na(accession)")
show(msnid)

```


## Accession FDR


```{r protein_level_filter }
fst_dms_pth_2 <- remap_fasta_entry_names(fst_dms_pth, conv, "^.*\\|(.+?)(-\\d+)?\\|.*")
fst <- readAAStringSet(fst_dms_pth_2)
names(fst) <- sub("^([A-Z]P_\\d+\\.\\d+)\\s.*", "\\1", names(fst))

## peptide density and accession filter
msnid <- compute_num_peptides_per_1000aa(msnid, fst_dms_pth_2)
msnid <- filter_msgf_data_protein_level(msnid, 0.01)
show(msnid)

```


## Parsimonious inference

To reduce number of protein identifications, we use a parsimonious inference algorithm described in 
[(Zhang et al. 2007)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2810678/).

```{r parsimonious_inference}
msnid <- infer_parsimonious_accessions(msnid, unique_only=FALSE)
show(msnid)

```


## Remove Decoys

```{r}
msnid <- apply_filter(msnid, "!isDecoy")
show(msnid)

save(msnid, file=file.path(data_folder, "msgfData_filtered.RData"))

```


## Process MASIC data

Output from the MASIC software is read from DMS, then filtered by interference score.


```{r read_masic}
masic_data_path <- file.path(data_folder, "masicData_original.RData")

if (file.exists(masic_data_path)) {
  load(masic_data_path)
} else {
  masic_data <- read_masic_data_from_DMS(data_package_num,
                                       interference_score = TRUE)
  save(masic_data, file=masic_data_path)
}

nrow(masic_data)
```


```{r filter_masic}
masic_data <- filter_masic_data(masic_data,
                                interference_score_threshold = 0.5,
                                s2n_threshold = 0)
save(masic_data, file=file.path(data_folder, "masicData_filtered.RData"))

nrow(masic_data)
```


## Create crosstab

The quantitative crosstab combines MS/MS identifications with reporter ion intensities. 
Abundances are taken relative to the reference channel and then log-transformed.


```{r create_global_gene_crosstab}
aggregation_level <- c("accession")
quant_cross_tab <- create_crosstab(msnid, 
                                   masic_data, 
                                   aggregation_level, 
                                   fractions, samples, references)
dim(quant_cross_tab)

write.table(quant_cross_tab, file=file.path(data_folder, paste(crosstab_prefix, "crosstab_global_gene_original.txt", sep="_")),
            quote=F, sep="\t")
```



