---
title: "Ex18 Combining datasets (Global)"
output: html_document
---




```{r include=FALSE}
library(MSnSet.utils)
library(dplyr)
library(ggplot2)
library(readxl)

source("../util/synapseUtil.R")
syn <- synapseLogin()

```


## Ex 18 data setup


```{r}
phospho_crosstab <- read.table("data/phospho_data/ptrc_ex18_crosstab_phospho_siteID_original.txt", sep = "\t")
colnames(phospho_crosstab) <- sub("^X", "", colnames(phospho_crosstab))

meta <- read_excel("../proteomics/data/Exp18_metadata_01-14-2021.xlsx") %>%
  filter(!is.na(Index)) %>%
  mutate(Index = as.character(Index),
         Plex = paste0("Ex18_", as.character(Plex)),
         Sample = paste0("Ex18_Sample_", Index)) %>% as.data.frame()
rownames(meta) <- meta$Index
meta <- meta[colnames(phospho_crosstab), ]
colnames(phospho_crosstab) <- meta$Sample
rownames(meta) <- meta$Sample

m_ex18 <- MSnSet(exprs = as.matrix(phospho_crosstab), pData = meta)
plot_pca(m_ex18, "Plex")

```


### Missing data


```{r}
peptide_missing <- rowSums(is.na(phospho_crosstab))
hist(peptide_missing)

```

## Ex 12 data setup

```{r}
ex_12 <- read.table(syn$get("syn22136309")$path, sep = "\t", header = T)
ex_12_peptide_df <- ex_12 %>% select(Site = site, Gene, Protein, Peptide) %>% unique() %>%
  group_by(Site) %>%
  mutate(total = n()) %>%
  ungroup()
ex_12 <- ex_12 %>%
  select(-Protein, -Gene, -Peptide) %>%
  unique()
ex_12_avg <- ex_12 %>%
  tidyr::pivot_longer(-site, names_to = "Sample", values_to = "value") %>%
  filter(!is.na(value),
         site != "") %>%
  tidyr::pivot_wider(names_from = "Sample", values_from = "value", values_fn = "mean") %>%
  as.data.frame()
ex_12_avg <- ex_12_avg[, colnames(ex_12)]
ex_12_rownames <- ex_12_avg$site
ex_12_avg <- ex_12_avg %>% select(-site)
rownames(ex_12_avg) <- ex_12_rownames

## Plex 1 and 2 contain MOLM14, while Plex 3 and 4 contain MV411. So select only 1 and 2 here.
meta_ex12 <- read_excel("data/LabelingMetadata_exp12.xlsx", sheet = "Sheet1") %>%
  mutate(sample = paste0("Sample_", `Sample ID`),
         Sample = paste0("Ex12_", sample),
         Plex = paste0("Ex12_", as.character(Plex)),
         state = sub("(^[A-Z,a-z]+) .*$", "\\1", `Sample Info`),
         State = case_when(state == "parental" ~ "Parental",
                           TRUE ~ state),
         Treatment = case_when(State == "Parental" ~ "none",
                               TRUE ~ "G"),
         Ligand = case_when(State == "Parental" ~ "none",
                            TRUE ~ sub("^.*, ([A-Z,0-9]+)$", "\\1", `Sample Info`))) %>%
  head(40) %>% as.data.frame()
rownames(meta_ex12) <- meta_ex12$sample
meta_ex12 <- meta_ex12[colnames(ex_12_avg), ]
colnames(ex_12_avg) <- meta_ex12$Sample
rownames(meta_ex12) <- meta_ex12$Sample

ex_12_avg <- ex_12_avg[, meta_ex12 %>% filter(Plex %in% c("Ex12_1", "Ex12_2")) %>% pull(Sample)]

m_ex12 <- MSnSet(exprs = as.matrix(ex_12_avg), 
                 pData = meta_ex12[colnames(ex_12_avg), ])

plot_pca(m_ex12, "Plex")

```

### Missing data

```{r}
phospho_ex12_missing <- rowSums(is.na(ex_12_avg))
hist(phospho_ex12_missing)

```




Now we take the sites which appear in each experiment and intersect to find a common
set of identified sites.


```{r}
common_identified_sites <- intersect(rownames(ex_12_avg), rownames(phospho_crosstab))

phospho_crosstab <- phospho_crosstab[common_identified_sites, ]
phospho_crosstab_ex12 <- ex_12_avg[common_identified_sites, ]

meta_combined <- rbind(meta %>% select(Sample, Plex, State, Treatment, Ligand), 
                       meta_ex12 %>% filter(Plex %in% c("Ex12_1", "Ex12_2")) %>% select(Sample, Plex, State, Treatment, Ligand)) %>%
  mutate(Exp = sub("_[0-9]$", "", Plex))

```


## Raw data stats


```{r}
phospho_crosstab <- sweep(phospho_crosstab, 2, apply(phospho_crosstab, 2, median, na.rm = T), FUN = '-')
phospho_crosstab_ex12 <- sweep(phospho_crosstab_ex12, 2, apply(phospho_crosstab_ex12, 2, median, na.rm = T), FUN = '-')

```

The distribtuion in each dataset is varied, even though we have similar cell lines.
So, we will normalize using a single mean and standard deviation for each dataset, 
independent of peptides, batches, etc, to better compare the two datasets. 


```{r}
hist(as.matrix(phospho_crosstab), main = "Ex18 data distribution")
hist(as.matrix(phospho_crosstab_ex12), main = "Ex12 data distribution")

```


```{r}
ex18_median <- median(as.matrix(phospho_crosstab), na.rm = T)
ex18_sd <- sd(as.matrix(phospho_crosstab), na.rm = T)
ex12_median <- median(as.matrix(phospho_crosstab_ex12), na.rm = T)
ex12_sd <- sd(as.matrix(phospho_crosstab_ex12), na.rm = T)

crosstab_ex18 <- phospho_crosstab
crosstab_ex12 <- phospho_crosstab_ex12

crosstab_ex18_norm <- (phospho_crosstab - ex18_median)/ex18_sd
crosstab_ex12_norm <- (phospho_crosstab_ex12 - ex12_median)/ex12_sd

```


## Combining the data 


### Normalized first, combined then corrected.


```{r}
crosstab_combined_norm <- cbind(crosstab_ex18_norm, crosstab_ex12_norm)
crosstab_combined_norm <- sweep(crosstab_combined_norm, 1, 
                                apply(crosstab_combined_norm, 1, median, na.rm = T), FUN = '-')

## Missing data 
combined_missing <- rowSums(is.na(crosstab_combined_norm))
crosstab_combined_norm <- crosstab_combined_norm[combined_missing < 30, ]

m <- MSnSet(exprs = as.matrix(crosstab_combined_norm), pData = meta_combined)
m_corrected <- correct_batch_effect_NA(m, batch_name = "Plex", par.prior = TRUE)
# m_corrected_rollup <- rrollup(m_corrected, rollBy = "Gene", rollFun = '-')

plot_pca(m, "Plex")
# plot_pca(m, "Plex", components = c(3,4))
plot_pca(m_corrected, "Plex")
# plot_pca(m_corrected, "Plex", components = c(3,4))
plot_pca(m_corrected, "Treatment")
# plot_pca(m_corrected, "Treatment", components = c(3,4))


```


### Combined (not normalized) then corrected


```{r}
crosstab_combined <- cbind(crosstab_ex18, crosstab_ex12)
crosstab_combined <- sweep(crosstab_combined, 1, 
                           apply(crosstab_combined, 1, median, na.rm = T), FUN = '-')

## Missing data 
combined_missing <- rowSums(is.na(crosstab_combined))
crosstab_combined <- crosstab_combined[combined_missing < 30, ]

m <- MSnSet(exprs = as.matrix(crosstab_combined), pData = meta_combined)
m_corrected <- correct_batch_effect_NA(m, batch_name = "Plex", par.prior = TRUE)
# m_corrected_rollup <- rrollup(m_corrected, rollBy = "Gene", rollFun = '-')

plot_pca(m, "Plex")
# plot_pca(m, "Plex", components = c(3,4))
plot_pca(m_corrected, "Plex")
# plot_pca(m_corrected, "Plex", components = c(3,4))
plot_pca(m_corrected, "Treatment")
# plot_pca(m_corrected, "Treatment", components = c(3,4))

## Using this one!
write.table(exprs(m_corrected), sep = "\t", file = "data/phospho_data/exp_12_18_combined_phospho_corrected.txt", quote = F)
# write.table(pData(m_corrected_rollup), sep = "\t", file = "data/exp_12_18_combined_metadata.txt", quote = F)

```

```{r}


```
